---
title: "IODS-course final assignment"
author: "Elina Turunen"
date: 1.3.2017
e-mail: elina.turunen@helsinki.fi
output: html_document
code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Happiness is nowadays proposed as a measure of well-being. It has been been questioned a lot for different reasons related to for example method and normativity. World Happiness Report (2016) examines happiness as a measure of well-being. Here some other than the more traditionally used variables of well-being will be looked at and the data of the report will be used in this exercise.

#### Research questions 

First the basic setting will be explored 
- Is there correlation between the happiness score and other variables?

The happiness score is quite vague. Autonomy of people seems more interesting in a sense that it has to do with rights, not affects or emotions. They are not unnecessary, but hard to propose as a legitimate policy targets.
- Is there correlation between freedom and other variables? (Are other variables contributing to freedom?)

Some aspects of data might be more effective than others, perhaps not happiness nor freedom, so it is good to explore. 
- What explains variance in the data, which variables are central in dividing the data?

TÄMÄ VIELÄ ENNEN EDELLISTÄ?
It is also interesting to see if there is some point where most countries differ in relation to freedom.
- What is the cut point of the freedom variable? MIKÄ TÄMÄ OLISI ?
VAI TÄMÄ
It is also interesting to see if there is some division of data, a number of clusters the data will divide at best.
- What is the optimal number of clusters the data will divide itself?


#### Data

The data originates from the World Happiness Report and it is downloaded from kaggle.com. The first such report was published in 2012 in support of the High Level Meeting at the United Nations on happiness and well-being. 2016 update is edited by John Helliwell, Richard Layard and Jeffrey Sachs. The report is published by The Sustainable Development Solutions Network (SDSN) that was commissioned by UN Secretary-General Ban Ki-moon. 

The report is a survey of the state of global happiness. It examines happiness and subjective well-being as measures human development. It was done also for policy use. Survey questions originate from the Gallup World Poll that was carried out by Gallup Inc. It provided data for the study. Some measures are created with data from the World Bank, OECD and WHO.


The variables of the data are as follows: *NÄMÄ BULLET POINTS

Happiness scores are based on answers to the life evaluation question of rating current lives on a scale from 0 (worst) to 10 (best), that is the so-called Cantril ladder. Happiness rank is the happiness score put in order. The happiness score is intended to represent people's happiness. *It is though not so clear what different people mean by that and mean by that at different times. 

GDP per capita is gross domestic product per capita in purchasing power parity, PPP, at dollar prices. The equation uses the natural log of GDP per capita. Healthy life expectancy is healthy life expectancy at birth. These variables represent state of economy and general level of health sector of countries.

Freedom is satisfaction with freedom to choose what to do with ones life (national average of binary responses). It portrays chances and limitations to act. Family variable is supposed to describe social support: having relatives or friends 
one can count on when help is needed (national average of binary responses). Trust is national average of responses to questions is corruption widespread throughout the government and within businesses (average of binary responses). Low values refer to corruption and high values to absence of corruption. *It can be asked whether the perceptions are trustworthy - are people having a rosy picture of their surroundings. For example Finland has been portrayed as a country of low corruption, but it is also known for example that corruption is not rare in construction business. 

Generosity is the residual of regressing national average of responses to question 
of donating money to a charity in the past month (binary) on GDP per capita. It is supposed to describe people's will to help others. Dystopia residual variable is a hypothetical country that has values equal to world’s lowest national averages for each of the six factors other than happiness. It was used as a benchmark against which countries can be compared in terms of six (key) variables and letting sub-bars (of these variables) to be of positive width in the report.

In the report 2016 update focus on inequality was new. Unfortunately some variables were not available in kaggle, for example the gini of household income and and gini index. They would have been important variables for the exercise.


#### Data wrangling 

Before analysis I changed column names to shorter descriptive names. I excluded variables happiness rank, region, dystopia and confidence intervals. I changed country column to name the rows and excluded the country column. (2)

Script is to be found here: LINK 


#### Explorations of the variables

First I looked at the distributions of the variables by drawing a histogram of each variable and looking at the summary of the variables. Then I looked at the relations of the variables by drawing ggpairs.

```{r, message=FALSE}
library(ggplot2); library(GGally); library(tidyr); library(dplyr)

happiness2 <- read.table("happiness2.csv", sep = ",", header=TRUE)

gather(happiness2) %>% ggplot(aes(value)) + geom_histogram() + facet_wrap("key", scales = "free") + ggtitle("Distributions of the variables")

summary(happiness2)

p <- ggpairs(happiness2, mapping = aes(), lower = list(combo = wrap("facethist", bins = 20))) + ggtitle("Correlations of the variables")
p
```

Most of the variables are binary of which national average has been taken. They are not distributed very evenly. The variables describe differences between countries in certain things. There are a few countries where there is perception of absence of corruption, the perception being more often that corruption is widespread in the government and within businesses. Social support is leaning towards higher social support. It is a variable not necessarily dependent on country's economy. GDP per capita, healthy life expectancy and freedom are also leaning to right that is higher GDP, life expectancy and freedom, GDP being a bit more varied. The generosity variable leans to lower generosity. Only a few countries have higher generosity. The happiness score is relatively evenly distributed if one can say so. So there are differences in rating life happiness. 

It is interesting that distributions of corruption and freedom reach only maximum values 0.6 for freedom and 0.5 for corruption.

GDP per capita, social support and healthy life expectancy have strong correlation with the happiness score, and quite strong for it and freedom. So it seems that there is correlation between the happiness score and other variables, which is no surprise since the data is very much planned and, I suppose, tested beforehand. Also freedom variable seems to be positively correlated with other variables. Strong correlation can be found between GDP per capita and healthy life expectancy and quite strong for social support; quite strong for social support and healthy life expectancy; and freedom and corruption.

(*What explains variance in the data, which variables are central in dividing the data? -> which seem to have variance ?
corruption, generosity, social support ? have ohut häntä
vai tuleeko varianssi tasaisemmin jakautuneista like the happiness score?)

(*Based on the distributions, it seems that the cut point of the freedom variable could be around 0.35?)


#### Methods

I will first do multiple regression analysis to see are other variables correlated with the happiness score, it being the dependent variable. Then I will then do the same for the freedom variable, so that x y ja z (not happiness?) being explaining variables and freedom being dependent.

I will then see what explains variance in the data by using the principal component analysis with which I can see which components capture the maximum amount of variance, as well as correlation between variables and components. 

Then I will (OR BEFORE?) look at the cut point of the freedom variable with the k-means OR k-means: determine k. That will tell what is the optimal number of clusters to divide the data. The purpose is to see what different (country?) groups there are in relation to freedom. With that we can perhaps find what is important affecting there.


#### Results  
Presentation of the results of your analysis including visualizations and summaries and a thorough interpretation of the results including a validation analysis of the method.

First I did multiple regression with happiness score as dependent variable. 

```{r}
m <- lm(happ.score ~ gdpcapita + socialsupport + hlifeexp + freedom + corruption + generosity, data = happiness2)

summary(m)

OR <- coef(m) %>% exp
CI <- confint(m) %>% exp
cbind(OR, CI)
```

In the first regression all other variables than generosity are statistically significant. When GDP per capita grows by one, happiness score grows by 0.7214. When social support grows by one, happiness score grows by 1.2298. When healthy life expectancy grows by one, happiness score grows by 1.4364. When freedom grows by one, hpapiness score grows by 1.5139. When absence of corruption grows by one, happiness score grows by 0.9189.

These variables (the model) explain 79 % of the happiness score.

Some of the confidence intervals are wide, especially for healthy life expectancy and freedom. **They are perhaps not the most powerful explanators for happiness. BUT THEY AFFECT SOMETIMES A LOT?

ODDS RATIOS...


Then I did multiple regression with freedom as dependent variable.

```{r}
m <- lm(freedom ~ gdpcapita + socialsupport + hlifeexp + corruption + generosity, data = happiness2)

summary(m)
```

In the second regression... 

So although...


KUVAT ?
CAPTIONS


# including a validation analysis of the method
Linear multiple regression is suitable method for the question, because...

# PCA

# interpretation of the results 
# including a validation analysis of the method


CAPTIONS


# k-means tai k-means: 

# interpretation of the results 
# including a validation analysis of the method


CAPTIONS


# Conclusions and discussion KRITIIKKIÄ mutta ehkä jo aiemmin perustelemaan valintoja JOS EI AIEMMIN NINI TÄHÄN VALIDATION JA happiness MITTAAMINEN YLIPÄÄnSÄ

(2)


# An ‘abstract’ at the beginning of the page with a summary of your analysis
(2) TIIVIS

# optional stuff




