---
title: "IODS-course final assignment"
author: "Elina Turunen"
date: 1.3.2017
e-mail: elina.turunen@helsinki.fi
output: html_document
code_folding: hide
---

Happiness is nowadays proposed as a measure of well-being. It has been been questioned a lot for different reasons related to for example method and normativity. World Happiness Report (2016) examines happiness as a measure of well-being. Here some other than the more traditionally used variables of well-being will be looked at and the data of the report will be used in this exercise.

#### Research questions 

First the basic setting will be explored 
- Is there correlation between the happiness score and other variables?

The happiness score is quite vague. Autonomy of people seems more interesting in a sense that it has to do with rights, not affects or emotions. They are not unnecessary, but hard to propose as a legitimate policy targets.
- Is there correlation between freedom and other variables? (Are other variables contributing to freedom?)

Some aspects of data might be more effective than others, perhaps not happiness nor freedom, so it is good to explore. 
- What explains variance in the data, which variables are central in dividing the data?

TÄMÄ VIELÄ ENNEN EDELLISTÄ?
It is also interesting to see if there is some point where most countries differ in relation to freedom.
- What is the cut point of the freedom variable? MIKÄ TÄMÄ OLISI ?
VAI TÄMÄ
It is also interesting to see if there is some division of data, a number of clusters the data will divide at best.
- What is the optimal number of clusters the data will divide itself?


#### Data

The data originates from the World Happiness Report and it is downloaded from kaggle.com. The first such report was published in 2012 in support of the High Level Meeting at the United Nations on happiness and well-being. 2016 update is edited by John Helliwell, Richard Layard and Jeffrey Sachs. The report is published by The Sustainable Development Solutions Network (SDSN) that was commissioned by UN Secretary-General Ban Ki-moon. 

The report is a survey of the state of global happiness. It examines happiness and subjective well-being as measures human development. It was done also for policy use. Survey questions originate from the Gallup World Poll that was carried out by Gallup Inc. It provided data for the study. Some measures are created with data from the World Bank, OECD and WHO.


The variables of the data are as follows: *BULLET POINTS

Happiness scores are based on answers to the life evaluation question of rating current lives on a scale from 0 (worst) to 10 (best), that is the so-called Cantril ladder. Happiness rank is the happiness score put in order. The happiness score is intended to represent people's happiness. *It is though not so clear what different people mean by that and mean by that at different times. 

GDP per capita is gross domestic product per capita in purchasing power parity, PPP, at dollar prices. The equation uses the natural log of GDP per capita. Healthy life expectancy is healthy life expectancy at birth. These variables represent state of economy and general level of health sector of countries.

Freedom is satisfaction with freedom to choose what to do with ones life (national average of binary responses). It portrays chances and limitations to act. Family variable is supposed to describe social support: having relatives or friends 
one can count on when help is needed (national average of binary responses). Trust is national average of responses to questions is corruption widespread throughout the government and within businesses (average of binary responses). Low values refer to corruption and high values to absence of corruption. *It can be asked whether the perceptions are trustworthy - are people having a rosy picture of their surroundings. For example Finland has been portrayed as a country of low corruption, but it is also known for example that corruption is not rare in construction business. 

Generosity is the residual of regressing national average of responses to question 
of donating money to a charity in the past month (binary) on GDP per capita. It is supposed to describe people's will to help others. Dystopia residual variable is a hypothetical country that has values equal to world’s lowest national averages for each of the six factors other than happiness. It was used as a benchmark against which countries can be compared in terms of six (key) variables and letting sub-bars (of these variables) to be of positive width in the report.

In the report 2016 update focus on inequality was new. Unfortunately some variables were not available in kaggle, for example the gini of household income and and gini index. They would have been important variables for the exercise.


#### Data wrangling 

Before analysis I changed column names to shorter descriptive names. I excluded variables happiness rank, region, dystopia and confidence intervals. I changed country column to name the rows and excluded the country column. (2)

Script is to be found here: LINK 


#### Explorations of the variables

First I had to fix rownames, because they did not preserve. 

```{r}
happiness2 <- read.table("happiness2.csv", sep = ",", header=TRUE)
str(happiness2)
colnames(happiness2)[1] <- "country"
rownames(happiness2) <- happiness2$country
rownames(happiness2)
happiness2 <- dplyr::select(happiness2, -country)
```

Then I looked at the distributions of the variables by drawing a histogram of each variable and looking at the summary of the variables. Then I looked at the relations of the variables by drawing ggpairs.

```{r, message=FALSE}
library(ggplot2); library(GGally); library(tidyr); library(dplyr)

gather(happiness2) %>% ggplot(aes(value)) + geom_histogram() + facet_wrap("key", scales = "free") + ggtitle("Distributions of the variables")

summary(happiness2)

p <- ggpairs(happiness2, mapping = aes(), lower = list(combo = wrap("facethist", bins = 20))) + ggtitle("Correlations of the variables")
p
```

Most of the variables are binary of which national average has been taken. They are not distributed very evenly. The variables describe differences between countries in certain things. There are a few countries where there is perception of absence of corruption, the perception being more often that corruption is widespread in the government and within businesses. Social support is leaning towards higher social support. It is a variable not necessarily dependent on country's economy. GDP per capita, healthy life expectancy and freedom are also leaning to right that is higher GDP, life expectancy and freedom, GDP being a bit more varied. The generosity variable leans to lower generosity. Only a few countries have higher generosity. The happiness score is relatively evenly distributed if one can say so. So there are differences in rating life happiness. 

It is interesting that distributions of corruption and freedom reach only maximum values 0.6 for freedom and 0.5 for corruption.

GDP per capita, social support and healthy life expectancy have strong correlation with the happiness score, and quite strong for it and freedom. So it seems that there is correlation between the happiness score and other variables, which is no surprise since the data is very much planned and, I suppose, tested beforehand. Also freedom variable seems to be positively correlated with other variables. Strong correlation can be found between GDP per capita and healthy life expectancy and quite strong for social support; quite strong for social support and healthy life expectancy; and freedom and corruption.

(*What explains variance in the data, which variables are central in dividing the data? -> which seem to have variance ?
corruption, generosity, social support ? have ohut häntä
vai tuleeko varianssi tasaisemmin jakautuneista like the happiness score?)

(*Based on the distributions, it seems that the cut point of the freedom variable could be around 0.35?)


#### Methods

I will first do multiple regression analysis to see are other variables correlated with the happiness score, it being the dependent variable. Then I will then do the same for the freedom variable, so that x y ja z (not happiness?) being explaining variables and freedom being dependent.

I will then see what explains variance in the data by using the principal component analysis with which I can see which components capture the maximum amount of variance, as well as correlation between variables and components. 

Then I will (OR BEFORE?) look at the cut point of the freedom variable with the k-means OR k-means: determine k. That will tell what is the optimal number of clusters to divide the data. The purpose is to see what different (country?) groups there are in relation to freedom. With that we can perhaps find what is important affecting there.


#### Results  
Presentation of the results of your analysis including visualizations and summaries and a thorough interpretation of the results including a validation analysis of the method.

First I did multiple regression with happiness score as dependent variable. 

```{r}
m <- lm(happ.score ~ gdpcapita + socialsupport + hlifeexp + freedom + corruption + generosity, data = happiness2)

summary(m)

CI <- confint(m) %>% exp
cbind(CI)
```

In the first regression all other variables than generosity are statistically significant, which was expected due to previous use of the data. The correlation is positive in all cases. For example when GDP per capita grows by one, happiness score grows by 0.7214. (When social support grows by one, happiness score grows by 1.2298. When healthy life expectancy grows by one, happiness score grows by 1.4364. When freedom grows by one, happiness score grows by 1.5139. When absence of corruption grows by one, happiness score grows by 0.9189.) These variables explain 78 % of the variation of the happiness score that is a lot. 

I wonder why generosity is not significant in this model, is its relative effect compared to other variables low.

Some of the confidence intervals are wide, especially for healthy life expectancy and freedom. The value of happiness with freedom is between 2.1-9.8 with the probability of 95 %. (*They are perhaps not the most powerful explanators for happiness.) LUOTTAMUSVÄLIT POIS?

Healthy life expectancy, GDP per capita and absence of corruption refer to the organization of the state. Social support is not necessarily dependent on that, it is possible to interpret the question concerning neighborhood or personal situations. Freedom is also a bit loose concept, it is possible to understand the question in many ways - freedom to choose what to do in relation to law and human rights or in relation to education system and social policy or in relation to personal situations. For example one could answer having high level of freedom in society in general even if one had personal hindrances such as need for rehabilitation that was not properly received. 

Research around happiness is also related to sustainability goals concerning also environmental sustainability. In that sense I understand that it is relevant to examine subjective well-being and not just objective and material aspects. But freedom seems to me a bit more concrete or material goal than happiness. You can be happy, but not free or unhappy, but free. Freedom at least relates to some rights and possibilities. There are many difficulties that lie in subjective questionings. Some connect happiness studies to positive psychology and compulsion of consumer culture to be happy. Therefore we can ask in what sense it is contributing to sustainability.

So I'm interested in which variables have connection to freedom, are they forming the material chances and limitations behind it. I did multiple regression with freedom as dependent variable.

```{r}
m <- lm(freedom ~ gdpcapita + socialsupport + hlifeexp + corruption + generosity, data = happiness2)

summary(m)

CI <- confint(m) %>% exp
cbind(CI)
```

In the second regression only social support, absence of corruption and generosity are statistically significant. There is positive correlation between these variables and freedom. For social support and generosity the correlation is not so strong. When absence of corruption grows by one, freedom grows by 0.46. It is comprehensible that absence of corruption in societies is a building block of freedom to do what one wants with same rules for everyone. But why is GDP per capita and healthy life expectancy not significant here? Also the estimates are very low. I would have guessed these material indicators had more effect. Though the GDP per capita does not tell about the distribution of wealth in societies. Here gini would be better variable to see whether material conditions have to do with freedom. Here it seems that freedom had to do more with trust-related variables social support, corruption and generosity. 

The confidence intervals for the significant variables vary between around 0.2-0.5 so they are not supernarrow when considered in relation to the distributions of the variables. *LUOTTAMUSVÄLIT POIS?

Linear multiple regression is suitable method for the question, when we want to examine correlations of variables and the dependent variable is continuous. *PAIKKA


Next I want to see which variables seem to be most important with principal component analysis. I need to first standardize the variables and then perform the analysis. Then I take rounded percentages of variance captured by principal components and create objects for axis labels before drawing a biplot.

```{r message=FALSE}
library(MASS); library(scales)
happiness2_std <- scale(happiness2)
pca_happiness2_std <- prcomp(happiness2_std)
pca_happiness2_std
s_std <- summary(pca_happiness2_std)
s_std
pca_pr_std <- round(100*s_std$importance[2,], digits = 1)
pca_pr_std
pc_lab_std <- paste0(names(pca_pr_std), " (", pca_pr_std,"%)")
biplot(pca_happiness2_std, cex = c(0.5, 1.5), xlab = pc_lab_std[1], ylab = pc_lab_std[2], sub = "Principal components explaining variance in standardized happiness2 data")
```

The first principal component PC1 captures the maximum amount of variance from the features in the original data that is 53,3 %. This combination of variables capture most essential dimensions of the data. GDP per capita, healthy life expectancy, social support and happiness score contribute to PC1. The second principal component PC2 captures maximum amount of variability left. Freedom, corruption and generosity contribute to PC2. They explain 19,9 % of the variance.

So it seems that GDP per capita, healthy life expectancy, social support and happiness score explain most of the variance from the features in the happiness2 data. It means that they explain greater share of differences between countries. Freedom, corruption and generosity are not be so constituent as them in making or explaining differences and these trust-related variables not so essential in this sense.

In that sense happiness is a feature that we can find differences in. Differences in GDP per capita and healthy life expectancy seem quite expected, like between industrialised and developing countries. In the biplot we can see a high concentration of rich countries in the lower left corner. Not all x- and y-axis features are uncorrelated.

The angle between the variables on the x-axis is quite small so the variables seem to have quite high correlation with each other. The variables on the y-axis PC2 have a wider angle between them and a bit smaller correlation. The correlation between the features and the pc axis is stronger for the features on the x-axis than for them on the y-axis. Features on the y-axis seem to have smaller standard deviations based on the lengths of the arrows (at least when enlarging the picture) and we can see that also in the table. It means that there are greater differences between countries in relation to features on the x-axis.

I don't know how to validate the principal component analysis method. It seems to be working well compared to previous knowledge of the data.


From here on I would like to look at the...


```{r}
set.seed(123)
dist_eu <- dist(happiness2, method = "euclidean")
k_max <- 10
twcss <- sapply(1:k_max, function(k){kmeans(dist_eu, k)$tot.withinss})
plot(1:k_max, twcss, type='b')
km <-kmeans(dist_eu, centers = 2)
pairs(happiness2, col = km$cluster)
```



Then I want to divide freedom into two, based on the distributions point where it seems to start getting higher values. I decided to divide it at 0.35. 

```{r}
happiness2_f <- mutate(happiness2, high_freedom = freedom > 0.35)
g2 <- ggplot(data = happiness2_f, aes(x = high_freedom))
g2 + geom_bar() + ggtitle("Distribution of the high freedom variable")
```

It seems that greater share of values for freedom are high with the chosen limit. Then I do (this time logistic) regression again, without the happiness score, to see which variables have effect on the higher values of freedom.  

```{r}
m <- glm(high_freedom ~ gdpcapita + socialsupport + hlifeexp + corruption + generosity, data = happiness2_f, family = "binomial")
summary(m)
OR <- coef(m) %>% exp
CI <- confint(m) %>% exp
cbind(OR, CI)
```



Social support, corruption and generosity are statistically significant.


a thorough interpretation of the results 
including a validation analysis of the method
CAPTIONS

PALJON VAIHTELUA SOCIAL SUPPORTISSA ?
MITEN MENEE ? onko loogisen tuntuinen ?

menetelmien yhdistäminen
lda + regressio


# Conclusions and discussion 

-regressiot
GDP per capita, healthy life expectancy, social support and happiness score explain most of the variance from the features in the happiness2 data and thereby explain greater share of differences between countries. There are also a little bit greater differences between countries in relation to these features than others. Trust-related variables freedom, corruption and generosity are not be so constituent as them in making or explaining differences.
-



# An ‘abstract’ at the beginning of the page with a summary of your analysis
# optional stuff




